---
title: "Vignette of ggplot2"
author: "Alex Khaykin"
date: "2023-03-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# INTRODUCTION

ggplot2 was the first of Hadley Wickham's tidy packages and was intended to simplify and streamline the appearance of R graphics. In this vignette, we will walk through key plots in ggplot2 using the 'congress_age' dataset from fivethirtyeight and best tidy practices.

## First, load the fivethirtyeight package and the congress_age dataset:
```{r}
# install.packages("fivethirtyeight")
library(fivethirtyeight)
data("congress_age")
str(congress_age)
```

Load the tidyverse and ggplot2 packages:
```{r}
library(ggplot2)
library(tidyverse)
library(dplyr)
```

## What are the top 10 most common first names of congresspeople?
First, we need to use the dplyr package to count and then sort the number of first names in the dataset.
```{r}
first_names <- congress_age %>% 
  group_by(firstname) %>% 
  count(firstname) %>% 
  arrange(desc(n))
head(first_names)
```

This uses the **group_by** function to group the congresspeople by their first names so they we can **count** them, and then we **arrange** them in descending (**desc**) order by the count (**n**) we generated.

#### Barplot with geom_bar
Barplots with geom_bar are a very quick way to look at summary data like counts. Although geom_bar will do the counting for you, here I are passing a dataframe that has already been summarized in counts so I will use the **stat="identity"** parameter inside of geom_bar.
```{r}
first_names[1:10,] %>% 
  ggplot(aes(y = n, x = firstname)) +
  geom_bar(stat = "identity") +
  labs(x = "First Name", y = "Frequency")

```

Please note that x and y labels are added by using the **labs()** function. Unlike with the dplyr or tidyverse, ggplot requires + signs rather than a %>% to separate the statements. For all ggplots the aesthetic mapping **aes()** is vital as well as some form of geom statement. What is passed through the aesthetic determines what is on the x and y axis. 

For example, by default ggplot will place the x axis into alphabetical order rather than take the order provided by the table. To fix this I can pass an additional parameter **scale_x_discrete()**:

```{r}
level_order <- first_names[1:10, "firstname"] 
first_names[1:10,] %>% 
  ggplot(aes(y = n, x = firstname)) +
  geom_bar(stat = "identity") +
  labs(x = "First Name", y = "Frequency") +
  scale_x_discrete(limits = level_order$firstname) 
```
In the history of the US congress the frequency of the name John has outstripped other first name with William, James, and Robert not far behind.

## Does the median age of a congress persons differ by political party?
In order to answer this question I will use a box plot on the raw dataset without any tidy manipulation.

#### Box plot with geom_boxplot.
This will create a conventional box and whisker plot.
```{r}
congress_age %>% 
  ggplot(aes(x = party, y = age, colour = party)) +
  geom_boxplot() +
  labs(x = "Party", y = "Age in Years")
```
The median age is similar across all political parties, except the Libertarian(L) and the American Independent party(AL).

#### Violin plot with geom_violin.
Violin plots are an alternative to box plots that add more information than a box plot in terms of the underlying distribution of the data. I will create a violin plot with the same data as above to demonstrate the additional information that can be obtained.

```{r}
congress_age %>% 
  ggplot(aes(x = party, y = age, colour = party)) +
  geom_violin() +
  labs(x = "Party", y = "Age in Years")
```
From this plot I can that the distribution of ages is most similar between Democrats(D) and Republicans(R). The Libertarian group is not shown because there was only one in the dataset.

What if we would like to visualize this plot horizontally instead?
I can employ **coord_flip()** to flip the coordinates of the plot:

```{r}
congress_age %>% 
  ggplot(aes(x = party, y = age, colour = party)) +
  geom_violin() +
  labs(x = "Party", y = "Age in Years") +
  coord_flip()
```

#### Scaterplot with geom_point and geom_smooth.
Scaterplots in ggplot are accomplished with **geom_point()** function and one can choose to add an an optional regression line to the data using either **geom_smooth()** or **geom_abline()**. However geom_abline requires that you have already calculated the line of best fit or another line before ploting. Geom_smooth is the ggplot replacement for baseR abline().

```{r}
congress_age %>% 
  ggplot(aes(x = termstart, y = age)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Start of Term", y = "Age in Years") 
```


The scatterplot and regression line demonstrate that over time we are electing older people to congress. 


### Facet_wrap is one way to create multiple plot pannel within the same plot.
Lets use the above regression plot to test whether there is a difference between democrats and republicans at age at start of term. To create panels in a ggplot one can use either **facet_wrap()** or **facet_grid()**. Both functions perform similarly although facet_grid will create plots even for missing data where as facet_wrap will not. Here I used facet_wrap to demonstrate how the wraping works by adding "~z" where z is grouping variable.    

```{r}
congress_age %>% 
  filter(party == "D" | party == "R") %>% 
  ggplot(aes(x = termstart, y = age)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Start of Term", y = "Age in Years") +
  facet_wrap(~party)
```

