---
title: "Data 607 Tidyverse Create Assignment"
author: "Mohamed Hassan-El Serafi"
date: "2023-04-10"
output: html_document
---

For this assignment, I will use dplyr, stringr, and ggplot2 packages from the tidyverse library. The dataset used was obtained from Kaggle, detailing the amount of money American sports owners have contributed to political campaigns and organizations during the 2016, 2018 and 2020 election cycles.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
library(tidyverse)
library(DT)
library(data.table)
library(plyr)
library(janitor)
```


```{r}
df <- read.csv("https://raw.githubusercontent.com/moham6839/SPRING2023TIDYVERSE/main/sports-political-donations.csv", check.names = FALSE)
head(df)
```
```{r}
glimpse(df)
```

```{r}
df$Amount <- as.integer(str_replace_all(df$Amount, "\\$|,", "")) #gdd - combined your string replacements and class coercion into one line
```

### Which Political Party received the most contributors?

```{r}
#gdd - reordered/simplified your group_by/summarize workflow
new_df <- df %>%
    filter(Party == "Republican" | Party == "Democrat") %>%
    group_by(Party) %>% 
    dplyr::summarize(n = n()) %>% 
    arrange(desc(n))
new_df

```

```{r}
ggplot(new_df, aes(x=`n`, y=`Party`)) +
  geom_bar(stat="identity", position="dodge", width=0.7) + 
  labs(title="Number of Contributors for Each Political Party",
       x="Number of Contributors",
       y="Political Party") 
```


### Which sports owner donated the most?

```{r}
owner_df <- df %>%
  filter(Team != "Commissioner") %>%
  group_by(Owner, League) %>%
  dplyr::summarise(Amount = sum(Amount)) %>%
  arrange(desc(Amount)) 
DT::datatable(owner_df)
```

```{r}
#gdd - adjusted your bar plot so that the x-/y-values are in more standard positions (i.e. the x is a category, and the y is numeric), your bars are sorted, and then the coords are flipped so you still achieve what you wanted visually: a horizontal bar plot. This is better than setting the y-value as a category and the x-value as numeric and not doing the coord flip, as it's easy to get confused when you do it that way. Also updated your Amounts since you wanted them to represent millions of dollars, not dollars.
ggplot(owner_df[tail(order(owner_df$Amount), 10), ],
       aes(x = reorder(Owner, Amount / 1000000),
           y = Amount / 1000000, fill = League)) +
    geom_bar(stat="identity", position="dodge", width=0.7) + 
    labs(title="American Sports Owners That Politically Contributed the Most Money",
       x="Name of Sports Owners",
       y="Amount Contributed (in millions $)") +
    coord_flip()
```



### Which Leagues provided the most political donations?


```{r}
league_df <- df %>%
  group_by(League) %>%
  dplyr::summarise(Amount = sum(Amount)) %>%
  arrange(desc(Amount))
DT::datatable(league_df)
```

```{r}
ggplot(league_df, aes(x=`Amount`, y=`League`)) +
  geom_bar(stat="identity", position="dodge", width=0.7) + 
  labs(title="Number of Contributors for Each Political Party",
       x="Amount Contributed (in millions $)",
       y="Professional American Sports Leagues") 
```




### Which recipient received the most political donations?



```{r}
recipient_df <- df %>%
  group_by(Recipient, Amount) %>%
  dplyr::summarise(Amount = sum(Amount)) %>%
  #dplyr::count(Recipient) %>% 
  arrange(desc(Amount)) 
DT::datatable(recipient_df)
#setDT(recipient_df) [order(-n), .SD[1:10]]
```



```{r}
ggplot(recipient_df[tail(order(recipient_df$Amount), 10), ], aes(x=`Amount`, y=`Recipient`)) +
  geom_bar(stat="identity", position="dodge", width=0.7) + 
  labs(title="Top 10 Recipients That Received the Most Money",
       x="Amount Contributed (in millions $)",
       y="Recipients") 
```





### Which Political Party received the most amount of money in each Election Year?

```{r}
year_df <- df %>%
  filter(Party == "Republican" | Party == "Democrat") %>%
  group_by(`Election Year`, `Party`) %>%
  dplyr::summarise(Amount = sum(Amount)) %>%
  arrange(desc(Amount))
year_df
```

```{r}
ggplot(year_df, aes(`Election Year`, `Amount`, fill=`Party`)) +
  geom_bar(stat="identity", position="dodge", width=0.7) +
  labs(title="Total Amount Contributed to Each Political Party",
       x="Amount Contributed (in millions $)",
       y="Election Year")
```

## Tidyverse Extend (Nick Climaco)

### Who are the top league donors for each political party?

```{r}
top_donors_df <- df %>%
    filter(Party %in% c("Democrat", "Republican")) |> 
    group_by(Party, League) |> 
    dplyr::summarize(total_donation = sum(Amount))
```

```{r}
top_donors_df %>%
    slice(1:8) |> 
    ggplot(aes(x = reorder(League, total_donation), y = total_donation, fill = Party)) +
    geom_col(position = "dodge") +
    coord_flip() +
    facet_wrap(~ Party, scales = "free_y") +
    labs(y = "Total Donation Amount",
    title = "Top Donors to Political Parties by Sports League") +
    theme_minimal() +
    scale_y_log10() +
    theme(legend.position = "none", axis.title.y = element_blank()) 
```

