---
title: "Tidyverse Create Assignment"
author: "Kory Martin, extended by Taha Ahmad"
date: "2023-04-08"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)

path = 'https://github.com/alu-potato/DATA607/raw/main/Assignments/Week%209%20Project/titles.csv'

movies_df<- read_csv(path)

```

## Introduction

For this project, we are using the Netflix TV Shows and Movies dataset from [Kaggle](https://www.kaggle.com/datasets/victorsoeiro/netflix-tv-shows-and-movies). Per the description provided on the website, "[t]his data set was created to list all shows available on Netflix streaming, and analyze the data to find interesting facts. This data was acquired in July 2022 containing data available in the United States." This dataset is broken up into a credits and titles datasets. For the purposes of this exercise, we focused on using the titles subfile.

## dplyr

For this exercise, we are using the **dplyr** library in the tidyverse package to show a number of different capabailities that are possible via this library.


### Selection

Sometimes when you get data, you are only interested in a subset of the features that are available in the dataset. For this example, we are going to start by selecting a subset of the columns that we are interested in working with and then assigning them to a subset vector:

- title
- type
- release_year
- age_certification
- runtime
- imdb_score
- imdb_votes
- tmdb_popularity

```{r}

keep_cols <- c('title', 'release_year', 'age_certification', 'runtime', 'imdb_score', 
              'imdb_votes', 'tmdb_popularity')

```

Using our subset vector, we will use the **select** function in dplyr to select only these columns from the data frame. We wrap this vector of column names into the **all_of()** function that we will pass to the select function.

```{r}

movies_df %>% 
  select(all_of(keep_cols))

```
The **select** function can also be used in the opposite manner. Say we wanted to keep all of the columns except for the age_certification, we can do that by inserting a "-" in front of the column we want to remove.

```{r}

movies_df %>% 
  select(all_of(keep_cols),-age_certification)

```

Now that we've selected our data, we can use the **arrange** function to sort the data by the release year feature, in ascending order.
```{r}

movies_df %>% 
  select(all_of(keep_cols),-age_certification) %>%
  arrange(release_year)

```

If instead we wanted to sort the data in descending order, we can encase the variable that we are sorting on (in this case "release_year") inside of the **desc()** function.

```{r}

movies_df %>% 
  select(all_of(keep_cols),-age_certification) %>%
  arrange(desc(release_year))

```

Dplyr also provides us with the capability to sort by multiple columns at the same time. Since we have multiple movies with the same release_year it can be helpful to also sort on other attributes. Simply add other columns to the list that you would also want to sort by. Below we sort on imbd_score in addition to release_year.

```{r}

movies_df %>% 
  select(all_of(keep_cols),-age_certification) %>%
  arrange(desc(release_year), desc(imdb_score))

```

If we only wanted to preview the first n rows of the resulting dataset, we can use the **slice_head()** function to limit the output to just the first few rows in the data frame. We pass the number of rows we want to display into the function by assigning it to the function's **n** variable. For example, in the code below, we use n=5 to only show the first five rows of the output data.

```{r}

movies_df %>% 
  select(all_of(keep_cols),-age_certification) %>%
  arrange(desc(release_year), desc(imdb_score)) %>%
  slice_head(n=5)

```
If you simply wanted to select the top (or bottom) rows by value, dplyr also has the **slice_min()** (or **slice_max()**) functions. This allows us to simplify our code from above to get the same results.

```{r}

movies_df %>% 
  select(all_of(keep_cols),-age_certification) %>%
  slice_max(order_by = tibble(release_year,imdb_score),n=5)
```

### Filtering and Mutating

Filtering is an important feature of dplyr that allows us to select specific rows of data that meet certain criteria.

If we wanted to only show the rows in the dataframe that correspond to movies that were released earlier than the year 2000, we can do so using the library's **filter** function as so.
```{r}

movies_df %>% 
  select(all_of(keep_cols),-age_certification) %>%
  filter(release_year < 2000)

```

Additionally, **filter** allows for multiple conditions to take place at once. For example, we could find movies that were released in between the 1970s and the 2000s but not released during the 80s as those movies sucked.

```{r}

movies_df %>% 
  select(all_of(keep_cols),-age_certification) %>%
  filter(release_year < 2000 & release_year > 1970, !between(release_year,1980,1990))

```

In this example, this results in a dataframe showing the 10 films in the library that fit this criteria.

If instead, of filtering and only selecting those movies that fit our criteria, we wanted to label movies as being pre-2000 and post-2000. We can do this by using the library's **mutate** function and combining it with a conditional statement. For this example, we will use the **ifelse()** function to set the criteria that we will use for setting the value of a new variable "release_window" to "pre_2000" if the movie was released prior to the year 2000, and "post_2000" otherwise.

```{r}
movies_df %>%
  select(all_of(keep_cols),-age_certification) %>%
  mutate(release_window = ifelse(release_year < 2000, "pre_2000","post_2000" ))

```

### Grouping & Summarizing

Grouping a data set by a common category is a very useful feature that allows for the data to be viewed by common features and details. For this data we will begin by grouping the data by age_certification and collecting some general information on the data based on these groupings. We group the data using the **group_by()** function and passing the name of a single column or multiple columns that we want to use for our grouping.

With our data in groups, we can then combine this with the **summarize()** function to calculate summary metrics that represent the calculated summary measures that pertain to the set of rows pertaining to newly created groups. 

In this example, we group our films by the age_certification category, and calculate a number of metrics for the data.
```{r}

movies_df %>% 
  select(all_of(keep_cols)) %>%
  group_by(age_certification) %>%
  summarize(num_titles = n_distinct(title),
            total_runtime = sum(runtime),
            avg_runtime = mean(runtime),
            min_runtime = min(runtime),
            max_runtime = max(runtime),
            avg_imdb_score = mean(imdb_score, na.rm = TRUE),
            min_imdb_score = min(imdb_score, na.rm = TRUE),
            max_imdb_score = max(imdb_score, na.rm = TRUE))



```

Finally, we can take our new grouped data and sort it by one of the fields in the resulting dataframe. In this case, we will conclude by sorting the data in reverse order based on the total number of titles in the group (num_titles)

```{r}
movies_df %>% 
  select(all_of(keep_cols)) %>%
  group_by(age_certification) %>%
  summarize(num_titles = n_distinct(title),
            total_runtime = sum(runtime),
            avg_runtime = mean(runtime),
            min_runtime = min(runtime),
            max_runtime = max(runtime),
            avg_imdb_score = mean(imdb_score, na.rm = TRUE),
            min_imdb_score = min(imdb_score, na.rm = TRUE),
            max_imdb_score = max(imdb_score, na.rm = TRUE)) %>%
  arrange(desc(num_titles))

```

## Conclusion

**dplyr** is a very useful library in the tidyverse package that provides highly useful functions that make working with dataframes easier and very effective. You can access the documentation for the library [here](https://dplyr.tidyverse.org/) and download a "cheat sheet" describing the key funcions for the library, [here](https://github.com/rstudio/cheatsheets/blob/main/data-transformation.pdf)