---
title: "Tidyverse Create - Extended"
author: "Coco Donovan, Kory Martin"
date: "`r Sys.Date()`"
output:
  html_document: default
---

## Introduction:

For my tidyverse create assignment, I chose a data set containing roster
information for all NCAA Women's Basketball teams. I intend to use readr to read 
in my data, dplyr to manipulate my data and ggplot2 to display my analysis.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading/Installing Packages

```{r}
library(ggplot2)
library(dplyr)
library(readr)
```

## readr

### read_csv()

I used the read_csv() function to read in my csv of NCAA Women's Basketball 
roster info. Within read_csv() I used the col_types argument to change data types
of certain columns. Initially the columns that I changed inside read_csv() were
chr type, but to do meaningful analysis I needed to make them numerical, hence 
setting them to col_double() (which sets the data type of the column to a double).
This same logic can be applied to height_

```{r}
ncaa_wbb_rosters <- read_csv('https://raw.githubusercontent.com/Sports-Roster-Data/womens-college-basketball/main/wbb_rosters_2022_23.csv', col_types = list(total_inches = col_double(), height_ft = col_double(), height_in = col_double()))

ncaa_avg_height <- round(mean(ncaa_wbb_rosters$total_inches, na.rm = TRUE),2)
```

## dyplr

### distinct()

I am unsure what values the redshirt column could possibly take on, so I use a 
pipe and the distinct() function to highlight the possible values 'redshirt' may 
take on. I find that redshirt can either be 1, for "yes, a student athlete was 
redshirted," or 0, for "no, a student athlete was not redshirted."

```{r}
knitr::kable(ncaa_wbb_rosters %>%
  distinct(redshirt))
```

### select(), group_by(), summarize(), and arrange() 

Now, I wanted to get a glimpse of the teams with the tallest average height. To 
do this I used pipes. I selected the team variable and the total_inches variable.
Then I grouped by height using the group_by function. I then used the summarize 
function to provide the counts of players for players per team the corresponding 
average height per team.

```{r}
avg_heights <- ncaa_wbb_rosters %>%
  select('team','total_inches') %>%
  group_by(team) %>%
  summarize(number_of_players = n(), Avg_height = round(mean(total_inches),2)) %>%
  arrange(desc(Avg_height))

knitr::kable(head(avg_heights))
```

## ggplot2

### Plotting Average Heights per team

The primary package I used here was ggplot2, although I do start this code chunk 
off with some dplyr. I used top_n() so that I could get the stop six results 
for the teams with the highest average heights. Then I piped those top 6 results
into a bar chart. I made sure to change the order of the x-axis as the default is
to sort the items alphabetically, but I want the value to be sorted based 
height in a descending order. I then limited the x-axis to better range to show
the difference between the top average heights (using coord_cartesian()), and
finished off my visualization by displaying the numerical values on top of each 
bar using geom_text() and displaying a line to show the average height of an NCAA
women's basketball team for perspective.

```{r}
top_n(avg_heights, n=6, Avg_height) %>%
  ggplot(., mapping = aes(x=reorder(team, desc(Avg_height)),y=Avg_height, fill=team)) +
  geom_bar(stat='identity') +
  coord_cartesian(ylim = c(66,74)) +
  geom_hline(aes(yintercept=ncaa_avg_height,linetype = "NCAA Average Height")) +
  geom_text(aes(label = Avg_height), vjust = 1.5,
              position = position_dodge(width = 0.9))+
  xlab("Team") +
  ylab("Average Height (in)")
```

## Extending the Examples (by Kory Martin):

For this we will extend upon the examples presented, by looking at the number of international students that are attending the universities and gleaning some meaningful insights based on this.

We begin by using the **mutate** function in dplpyr to create a new column that will hold a value of 'international' if the student is not from the US, and a value of 'domestic' if they are from the US. In addition to the mutate function, we will use the **if_else** function to assign the value to the new column, based on whether or not the value of country_clean is 'USA' or not.

```{r}
(ncaa_wbb_roster_extended <- ncaa_wbb_rosters %>% 
  mutate(domestic_international = ifelse(country_clean == 'USA', 'domestic', 'international')))


```

We will then use **select** function combined with the **slice_sample** function to confirm that our new column has the expected values, by selecting a random sample of rows and looking at their country_clean and domestic_international values

```{r}
ncaa_wbb_roster_extended %>% select(country_clean, domestic_international) %>% slice_sample(n=15)
```

Next we will use the **group_by** function to generate a count of the number of domestic vs international students at each college. Furthermore, we will use the **summarize** function to create summary measures for the total number of domestic students, the total number of international students, and the total number of players on the team. Finally, based on these counts, we will then create an additional variable pct_international that will represent the pct of international students on the student rosters. 

```{r}
(ncaa_international_domestic <- ncaa_wbb_roster_extended %>%
  group_by(team) %>%
  summarize(num_domestic = sum(ifelse(domestic_international == 'domestic',1,0)),
            num_international = sum(ifelse(domestic_international == 'international',1,0)),
            num_players = n()) %>%
  mutate(pct_international = num_international/num_players))


```
We will use the **left_join** function to connect our data to the division in which the college belongs to
```{r}
(ncaa_international_domestic <- left_join(ncaa_international_domestic, ncaa_wbb_rosters %>% select(team, division) %>% distinct(), by=c("team"="team")))

```
Finally we will use the **group_by** and **summarize** functions, to determine the average pct of international students per team, across teams in each of the three divisions. 
```{r}
ncaa_international_domestic %>%
  group_by(division) %>%
  summarize(num_teams = n(),
            avg_pct_international = mean(pct_international))

```



## Conclusion:

As you can see Indiana has the tallest average height, followed by UConn, Michigan,
South Carolina, Rutgers and Stanford. As a huge women's basketball fan, one thing 
that stands out to me is that all but two of these teams have been nationally
ranked in the top 5 this season, and of the two that have not been ranked in the
top 5, Rutgers and Michigan, Michigan has been ranked in the top 20, pretty
consistently. Rutgers has been going through a rough patch in the absence of their
hall-of-fame coach C. Vivian Stringer and their roster only contains 8 players
(their average height may be inflated by the lack of players on the roster). However,
my first thought after looking at this would be that it seems that height plays
some component in basketball (stating the obvious).